---
title: "Clinical Predictors Trajectories Analysis -- ESS Illustration"
author: "Marissa Ashner"
date: "2026-04-28"
output:  
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

This script runs the analysis for the clinical predictor data with the Pain Interference trajectory outcomes for the ESS illustration.

Note: Without the data, this script will not run on its own. It is shared for the purposes of understanding and utilizing the methods involved with the ESS pipeline.

# Libraries and other setup

```{r echo = FALSE}
##### Packages Needed ##### 
library(dplyr)
library(caret)
library(glmnet)
library(tidyr)
library(mpath)
library(readxl)
library(stabs)
library(flextable)
library(purrr)
library(stringr)
library(here)
```

# Read in Data 

Data unavailable for public use. The rest of this script is run assuming an analysis dataset called "data" is loaded into the environment.

# Center/Scale and Factorize Covariates

```{r}
# Factor CMVIgG negative
data$CMVIgG_negative = as.factor(data$CMVIgG_negative)

# Center and Scale continuous clinical variables 
data = data %>% mutate(across(gait_speed:emotional_support_tscore & 
                                where(is.numeric),
                              ~ as.numeric(scale(.x))))
```

# Prep Datasets/ Covariates 

```{r}
# remove anyone with missing biomakers 
# select only the clinical predictors for the predictor set
data_ltc_pi = data %>% filter(!is.na(IFNy)) %>% 
  select(subject_id, pi_traj_cont, gait_speed:emotional_support_tscore, 
                                  -spef_diab)
```

# Make Function for Covariate Matrix 

```{r}
# function to make the design matrix for the predictors
# this is written for "standard LASSO" and not "group LASSO", meaning levels of variables will be considered by the penalization method separately and will not be tethered together                                                                                                                            
make_group_matrix <- function(
    data, # data frame
    response, # name of the outcome column
    # Factors
    na_as_level = TRUE, # creates a new level for the NAs
    na_level    = "(Missing)", # what to call the NA level
    # Numeric preprocessing
    center = TRUE, # center numeric variables
    scale  = TRUE, # scale numeric variables 
    indicator_suffix = "_miss" # suffix for missing indicator variable
    ) {
  
  # check to see if data is a data frame
  stopifnot(is.data.frame(data))
  
  # check to see if the response is provided as a character, length 1, and it's in the data frame
  stopifnot(is.character(response), length(response) == 1, response %in% names(data))

  # for the continuous missing indicators, we will impute the mean for the numeric portion of the variables
  impute_numeric <- "mean"

  # Work on a copy
  df <- data

  # Extract outcome and remove from predictors
  y <- df[[response]]
  df[[response]] <- NULL

  # Factors: coerce and (optionally) add explicit missing level
  for (nm in names(df)) {
    if (is.character(df[[nm]])) df[[nm]] <- factor(df[[nm]])
    if (is.factor(df[[nm]]) && isTRUE(na_as_level)) {
      df[[nm]] <- addNA(df[[nm]])
      lev <- levels(df[[nm]])
      lev[is.na(lev)] <- na_level
      levels(df[[nm]]) <- lev
    }
  }

  # Numeric main effects: indicators + impute + center/scale 
  cont_vars <- names(df)[vapply(df, function(z) is.numeric(z) && !is.factor(z), logical(1))]
  miss_indicator_cols <- list()
  preproc <- list(
    numeric = list(stats = list(),
                   method = impute_numeric,
                   center = center,
                   scale  = scale),
    factor  = list(na_as_level = na_as_level, na_level = na_level)
  )

  for (v in cont_vars) {
    xvec <- df[[v]]
    # 0/1 missingness indicator (kept even if there are no NAs; it's all zeros then)
    miss_indicator_cols[[v]] <- as.integer(is.na(xvec))

    # Impute
    fill <- switch(impute_numeric,
                   mean   = mean(xvec, na.rm = TRUE),
                   median = stats::median(xvec, na.rm = TRUE),
                   zero   = 0)
    x_imp <- xvec
    x_imp[is.na(x_imp)] <- fill

    # Center/scale
    mu  <- if (center) mean(x_imp) else 0
    sdv <- if (scale)  stats::sd(x_imp) else 1
    if (isTRUE(scale) && (is.na(sdv) || sdv == 0)) sdv <- 1

    df[[v]] <- if (!center && !scale) x_imp else (x_imp - mu) / sdv

    # Record stats for reproducibility
    preproc$numeric$stats[[v]] <- list(impute_value = fill, mean = mu, sd = sdv)
  }

  # Build model matrix
  tt <- terms(~ ., data = df)
  mf <- model.frame(tt, data = df, na.action = na.pass)

  X_full <- model.matrix(tt, data = mf)   # includes intercept if any
  assign_full <- attr(X_full, "assign")   # capture before any subsetting
  if (is.null(assign_full)) {
    stop("Internal error: attr(X_full, 'assign') is NULL; model.matrix did not supply 'assign'.")
  }

  # Drop intercept column and its assign entry if present
  has_intercept <- isTRUE(attr(tt, "intercept") == 1)
  if (has_intercept) {
    # Intercept is column 1 in model.matrix output
    X0 <- X_full[, -1, drop = FALSE]
    assign_vec <- assign_full[-1]
  } else {
    X0 <- X_full
    assign_vec <- assign_full
  }
  
  
  # Term labels (one per model term; often equals unique(assign_vec))
  term.labels <- attr(tt, "term.labels")

  # Start from base matrix and grouping
  X     <- X0
  group <- assign_vec
  
    # Append numeric missing indicators, grouped with their main-effect term
  if (length(miss_indicator_cols)) {
    for (v in names(miss_indicator_cols)) {
      # Find the *term id* for this variable in the ~ . design
      term_id <- match(v, term.labels)
      if (is.na(term_id)) next  # defensive; should not happen in ~ . design

      ind_name <- paste0(v, indicator_suffix)
      ind_vec  <- as.numeric(miss_indicator_cols[[v]])

      # Append as a *matrix* column to keep X as matrix
      ind_mat <- matrix(ind_vec, nrow = nrow(X), ncol = 1,
                        dimnames = list(NULL, ind_name))
      X <- cbind(X, ind_mat)

      # Append exactly one group id for the new column
      group <- c(group, term_id)
    }
  }
  
    # Ensure unique, clean column names
  colnames(X) <- make.unique(colnames(X), sep = "_dup")

  # Final integrity checks
  if (ncol(X) != length(group)) {
    stop(sprintf("Length mismatch: ncol(X) = %d but length(group) = %d",
                 ncol(X), length(group)))
  }


  list(X = X, y = y)
}


```




# Pain Interference

```{r}
# Create Model Matrix 
mm_pi <- make_group_matrix(
  data         = data_ltc_pi[,-1], # your data.frame; removed the ID column
  response  = "pi_traj_cont", # outcome column name in df
  center = FALSE, # already centered and scaled before subsetting
  scale = FALSE
)

## remove any columns with all zeros in this subset of the cohort
mm_pi$X <- mm_pi$X[, colSums(mm_pi$X != 0) > 0]


# Complementary Pairs Stability Selection 

## initialize results 
results_cpss_pi = list()

## loop over the number of active variables
for(q in seq(5, 25, 5)){
  # set seed for reproducibility 
  set.seed(339580)
  
  # run stability selection function
  temp = stabsel(x = mm_pi$X, # model matrix
               y = mm_pi$y, # repsonse
               fitfun = glmnet.lasso, # run LASSO
               q = q, # number of active variables 
               PFER = 2, # error rate; not being used for our purposes; choosing an arbitrary value
               B = 100 # number of complementary pairs
               )
  # grab the selection probabilities 
  results_cpss_pi[[paste0("q", q, "_cpss")]] = temp$max
}

## combine results  
combined_cpss_pi <- as.data.frame(do.call(cbind, results_cpss_pi))

combined_cpss_pi$predictor = rownames(combined_cpss_pi)

combined_cpss_pi <- combined_cpss_pi %>%
  mutate(q5_cpss_ranked = rank(-q5_cpss) %>% as.integer(), 
         q10_cpss_ranked = rank(-q10_cpss) %>% as.integer(),
         q15_cpss_ranked = rank(-q15_cpss) %>% as.integer(),
         q20_cpss_ranked = rank(-q20_cpss) %>% as.integer(),
         q25_cpss_ranked = rank(-q25_cpss) %>% as.integer())

# Bootstrap LASSO

# perform LOO cross-validation on full dataset to find corresponding lambda value for each number of active variables
cv_model <- cv.glmnet(mm_pi$X, # design matrix
                      mm_pi$y, # response
                      alpha = 1, # LASSO
                      type.measure = "mse", # type of error measure
                      nfolds = nrow(data_ltc_pi) # LOOCV
                      )

# pick the lambdas for each number of active variables 
zeros = cv_model$nzero
lambdas = c(cv_model$lambda[which.max(zeros[zeros <= 5])], 
            cv_model$lambda[which.max(zeros[zeros <= 10])], 
            cv_model$lambda[which.max(zeros[zeros <= 15])], 
            cv_model$lambda[which.max(zeros[zeros <= 20])], 
            cv_model$lambda[which.max(zeros[zeros <= 25])]
)

# initialize results list
results_boot_pi = list(q5_boot = c(), q10_boot = c(), q15_boot = c(), 
                           q20_boot = c(), q25_boot = c())

# Loop over number of resamples
for(B in 1:200){
  # set seed each iteration for reproducibility
  set.seed(B*12098)
  
  # sample bootstrap indices 
  indices = sample(nrow(data_ltc_pi), 
                   nrow(data_ltc_pi), 
                   replace = TRUE)
  boot_pi_X = mm_pi$X[indices,]
  boot_pi_y = mm_pi$y[indices]
  
  # run model on the resample
  final_model = glmnet(boot_pi_X, 
                     boot_pi_y,
                     alpha = 1,
                     lambda = lambdas)
  
  # get results for each penalty parameter
  for(i in 1:length(lambdas)){
      results_boot_pi[[i]] =
        c(results_boot_pi[[i]],
          rownames(coef(final_model, 
                        s = lambdas[i]))
          [coef(final_model, 
                s = lambdas[i])[, 1] != 0][-1])
  }
}

# organize results 
for(name in names(results_boot_pi)){
  results_boot_pi[[name]] = data.frame(
  table(results_boot_pi[[name]])/200
) %>% arrange(-Freq)
colnames(results_boot_pi[[name]]) <- c("predictor", name)
}

combined_boot_pi <- Reduce(function(x,y)
  full_join(x, y, by = "predictor"), results_boot_pi) %>% 
    mutate(across(where(is.numeric), ~ replace_na(., 0))) %>% 
  mutate(q5_boot_ranked = rank(-q5_boot, na.last = "keep") %>% as.integer(), 
         q10_boot_ranked = rank(-q10_boot, na.last = "keep") %>% as.integer(),
         q15_boot_ranked = rank(-q15_boot, na.last = "keep") %>% as.integer(),
         q20_boot_ranked = rank(-q20_boot, na.last = "keep") %>% as.integer(),
         q25_boot_ranked = rank(-q25_boot, na.last = "keep") %>% as.integer())

# Combining Results for CPSS and Bootstrap 
combined_pi = full_join(combined_cpss_pi, 
                            combined_boot_pi, 
                            by = "predictor")
combined_pi <- combined_pi %>% 
    mutate(across(where(is.numeric) & ends_with("ranked"), ~ replace_na(., 30))) %>% 
   mutate(across(where(is.numeric) & ends_with("boot"), ~ replace_na(., 0)))

top10_pi_binary = combined_pi %>% select(ends_with("ranked")) %>% apply(2, function(x) ifelse(x <= 10, 1, 0))
rownames(top10_pi_binary) = combined_pi$predictor

rowSums_top10_pi_binary = rowSums(top10_pi_binary) %>% sort(decreasing = TRUE) 

results_sp_pi = data.frame(predictor = names(rowSums_top10_pi_binary), 
           stabscore = unname(rowSums_top10_pi_binary)) %>% left_join(
             combined_pi, by = "predictor"
           ) %>% rowwise %>% mutate(
    median_sp = median(c_across(matches("^q(5|10|15|20|25)_(cpss|boot)$")))) %>% ungroup() %>% 
  arrange(-stabscore, -median_sp) %>% 
  filter(stabscore != 0) %>% 
  mutate(health_measure = "pi")
```


# save Top 10s

```{r}
# saveRDS(results_sp_pi, INSERT PATH NAME)
```

# Save model Matrices 

```{r}
# saveRDS(mm_pi, INSERT PATH NAME)
```


# Session information

```{r echo = FALSE, results = 'markup'}
sessionInfo()
```


