---
title: "Biomarkers Trajectories Analysis -- ESSF Illustration"
author: "Marissa Ashner"
date: "2026-04-28"
output:  
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

This script runs the analysis for the biomarker data with the functional trajectory outcomes for the ESS Illustration. This version of the script saves the selection probabilities.

Note: Without the data, this script will not run on its own. It is shared for the purposes of understanding and utilizing the methods involved with the ESS pipeline.

# Libraries and other setup

```{r echo = FALSE}
##### Packages Needed ##### 
library(dplyr)
library(caret)
library(glmnet)
library(tidyr)
library(mpath)
library(readxl)
library(stabs)
library(flextable)
library(purrr)
library(stringr)
library(here)
```

# Read in Data 

Data unavailable for public use. The rest of this script is run assuming an analysis dataset called "data" is loaded into the environment.


# Center/Scale and Factorize Covariates

Note: For the biomarkers in this dataset, they were centered and scaled before this step.

```{r}
# Factor CMVIgG negative
data$CMVIgG_negative = as.factor(data$CMVIgG_negative)
```

# Prep Datasets/ Covariates 

```{r}
### remove anyone with missing biomakers 
# select only the biomarkers for the predictor set
data_ltc_pi = data %>% filter(!is.na(IFNy)) %>% 
  select(subject_id, IFNy:CMVIgG_negative, pi_traj_cont)
```

# Pain Interference

```{r}
# Create Model Matrix 
mm_pi <- list(X = data_ltc_pi %>% select(-subject_id, -pi_traj_cont), 
                  y = data_ltc_pi$pi_traj_cont)

## make CMVIgG negative indicator numeric
mm_pi$X$CMVIgG_negative = as.numeric(mm_pi$X$CMVIgG_negative)-1

# remove predictors that are all zeros in this cohort
mm_pi$X <- mm_pi$X[, colSums(mm_pi$X != 0) > 0]

# Complementary Pairs Stability Selection 

## initialize results
results_cpss_pi = list()

## loop over the number of active variables
for(q in seq(5, 25, 5)){
  # set seed for reproducibility
  set.seed(339580)
  
  # run stability selection function
  temp = stabsel(x = mm_pi$X, # model matrix
               y = mm_pi$y, # response
               fitfun = glmnet.lasso, # run LASSO
               q = q, # number active variables
               PFER = 2, # error rate; not being used for our purposes; choosing an arbitrary value
               B = 100 # number of complementary pairs
               )
  # grab the selection probabilities 
  results_cpss_pi[[paste0("q", q, "_cpss")]] = temp$max
}

## combine results
combined_cpss_pi <- as.data.frame(do.call(cbind, results_cpss_pi))

combined_cpss_pi$predictor = rownames(combined_cpss_pi)

combined_cpss_pi <- combined_cpss_pi %>%
  mutate(q5_cpss_ranked = rank(-q5_cpss) %>% as.integer(), 
         q10_cpss_ranked = rank(-q10_cpss) %>% as.integer(),
         q15_cpss_ranked = rank(-q15_cpss) %>% as.integer(),
         q20_cpss_ranked = rank(-q20_cpss) %>% as.integer(),
         q25_cpss_ranked = rank(-q25_cpss) %>% as.integer())

# Bootstrap LASSO

# perform LOO cross-validation on full dataset to find corresponding lambda value for each number of active variables
cv_model <- cv.glmnet(mm_pi$X, # design matrix
                      mm_pi$y, # response
                      alpha = 1, # LASSO
                      type.measure = "mse", # type of error measure
                      nfolds = nrow(data_ltc_pi) # LOOCV
                      )

# pick the lambdas for each number of active variables
zeros = cv_model$nzero
lambdas = c(cv_model$lambda[which.max(zeros[zeros <= 5])], 
            cv_model$lambda[which.max(zeros[zeros <= 10])], 
            cv_model$lambda[which.max(zeros[zeros <= 15])], 
            cv_model$lambda[which.max(zeros[zeros <= 20])], 
            cv_model$lambda[which.max(zeros[zeros <= 25])]
)

# initialize results list
results_boot_pi = list(q5_boot = c(), q10_boot = c(), q15_boot = c(), 
                           q20_boot = c(), q25_boot = c())

# Loop over number of resamples
for(B in 1:200){
  # set seed each iteration for reproducibility
  set.seed(B*12098)
  
  # sample bootstrap indices 
  indices = sample(nrow(data_ltc_pi), 
                   nrow(data_ltc_pi), 
                   replace = TRUE)
  boot_pi_X = mm_pi$X[indices,]
  boot_pi_y = mm_pi$y[indices]
  
  # run model on the resample
  final_model = glmnet(boot_pi_X, 
                     boot_pi_y,
                     alpha = 1,
                     lambda = lambdas)
  
  # get results for each penalty parameter
  for(i in 1:length(lambdas)){
      results_boot_pi[[i]] = c(results_boot_pi[[i]],rownames(coef(final_model, 
                                            s = lambdas[i]))[coef(final_model, 
                                                                  s = lambdas[i])[, 1] != 0][-1])
  }
}

# organize results
for(name in names(results_boot_pi)){
  results_boot_pi[[name]] = data.frame(
  table(results_boot_pi[[name]])/200
) %>% arrange(-Freq)
colnames(results_boot_pi[[name]]) <- c("predictor", name)
}

combined_boot_pi <- Reduce(function(x,y)
  full_join(x, y, by = "predictor"), results_boot_pi) %>% 
    mutate(across(where(is.numeric), ~ replace_na(., 0))) %>% 
  mutate(q5_boot_ranked = rank(-q5_boot, na.last = "keep") %>% as.integer(), 
         q10_boot_ranked = rank(-q10_boot, na.last = "keep") %>% as.integer(),
         q15_boot_ranked = rank(-q15_boot, na.last = "keep") %>% as.integer(),
         q20_boot_ranked = rank(-q20_boot, na.last = "keep") %>% as.integer(),
         q25_boot_ranked = rank(-q25_boot, na.last = "keep") %>% as.integer())

# Combining Results for CPSS and Bootstrap
combined_cpss_pi$predictor = str_replace_all(combined_cpss_pi$predictor, 
                                             "`", "")
combined_pi = full_join(combined_cpss_pi, 
                            combined_boot_pi, 
                            by = "predictor")
combined_pi <- combined_pi %>% 
    mutate(across(where(is.numeric) & ends_with("ranked"), ~ replace_na(., 30))) %>% 
   mutate(across(where(is.numeric) & ends_with("boot"), ~ replace_na(., 0)))

top10_pi_binary = combined_pi %>% select(ends_with("ranked")) %>% apply(2, function(x) ifelse(x <= 10, 1, 0))
rownames(top10_pi_binary) = combined_pi$predictor

rowSums_top10_pi_binary = rowSums(top10_pi_binary) %>% sort(decreasing = TRUE) 

results_sp_pi = data.frame(predictor = names(rowSums_top10_pi_binary), 
           stabscore = unname(rowSums_top10_pi_binary)) %>% left_join(
             combined_pi, by = "predictor"
           ) %>% rowwise %>% mutate(
    median_sp = median(c_across(matches("^q(5|10|15|20|25)_(cpss|boot)$")))) %>% ungroup() %>% 
  arrange(-stabscore, -median_sp) %>% 
  filter(stabscore != 0) %>% 
  mutate(health_measure = "pi")
```

# save Top 10s

```{r}
# saveRDS(results_sp_pi, INSERT PATH NAME)
```

# Save model Matrices 

```{r}
# saveRDS(mm_pi, INSERT PATH NAME)
```


# Session information

```{r echo = FALSE, results = 'markup'}
sessionInfo()
```


