---
title: "ESSF Illustration -- Regressions and Outputs"
author: "Marissa Ashner"
date: "2026-04-28"
output:  
  html_document:
    toc: true
    toc_float:
        collapsed: true
        smooth_scroll: true
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Purpose

This script runs all regressions and creates forest plots for ESS Illustration; uses the stability scores to create tables and heat maps.

Note: Without the data, this script will not run on its own. It is shared for the purposes of understanding and utilizing the methods involved with the ESS pipeline.

# Libraries and other setup

```{r echo = FALSE}
##### Packages Needed ##### 
library(dplyr)
library(caret)
library(glmnet)
library(tidyr)
library(mpath)
library(readxl)
library(stabs)
library(here)
library(patchwork)
library(stringr)
library(flextable)
library(officer)

##### clear existing data and graphics #####  
rm(list=ls())
graphics.off()
```

# Read in Data 

Data unavailable for public use. The rest of this script is run assuming datasets of the stable sets and model matrices called "stable_sets_clin"/"mm_pi" (from program 02), "stable_sets_bm"/"mm_pi_bm" (from program 03), and "stable_sets_clbm"/"mm_pi_clbm" (from program 04) are loaded into the environment. Additionally, the analysis dataset loaded into programs 02, 03, and 04 is loaded into the environment as "analysis_dataset".

# Regressions 

## Unadjusted 

### PI Biomarkers

For each biomarker in the top 10 for pain interference, run an unadjusted regression model.

```{r}
# select the predictor names for the stable set
top10_pi_bm = stable_sets_bm %>%
  filter(health_measure == "pi") %>%
  select(predictor) %>% 
  unlist() %>% unname()

# Subset model matrix to those in the top 10
X_selected <- mm_pi_bm$X[, top10_pi_bm]

# new dataset -- filter the analysis dataset to those with biomarkers and choose appropriate predictors 
pi_data_bm = analysis_dataset %>% 
filter(!is.na(IFNy)) %>%
  select(subject_id, pi_traj_cont, illness6) %>%
  cbind(X_selected) %>% 
  na.omit()
```


```{r}
# running the unadjusted regression models

# Initialize a list to store coefficients
pi_unadj_coefficients_bm <- list()

# Loop through each covariate
for (predictor in top10_pi_bm) {
  formula <- as.formula(paste0("pi_traj_cont ~ `", predictor, "`"))
  
  # Fit the linear model
  model <- lm(formula, data = pi_data_bm)
  
  # Extract the coefficient for the covariate
  coef <- summary(model)$coefficients[-1, "Estimate"]

  # Store
  pi_unadj_coefficients_bm[[predictor]] <- coef
}

# Convert the list to a dataframe for better readability
pi_coefficients_df_bm <- data.frame(
  predictor = names(pi_unadj_coefficients_bm), 
  Coefficient = unlist(pi_unadj_coefficients_bm)
)
```

### PI Clinical 

```{r}
# select the predictor names for the stable set
top10_pi_clinical = stable_sets_clin %>%
  filter(health_measure == "pi") %>%
  select(predictor) %>% 
  unlist() %>% unname()

# Subset model matrix to those in the top 10
X_selected <- mm_pi$X[, top10_pi_clinical]

# new dataset -- filter the analysis dataset to those with biomarkers and choose appropriate predictors 
pi_data_clinical = analysis_dataset %>% 
filter(!is.na(IFNy)) %>% 
  select(subject_id, pi_traj_cont, illness6) %>% 
cbind(X_selected) %>% 
  na.omit()
```

```{r}
# running the unadjusted regression models

# Initialize a list to store coefficients
pi_unadj_coefficients_clinical <- list()

# Loop through each covariate
for (predictor in top10_pi_clinical) {
  formula <- as.formula(paste0("pi_traj_cont ~ `", predictor, "`"))
  
  # Fit the linear model
  model <- lm(formula, data = pi_data_clinical)
  
  # Extract the coefficient for the covariate
  coef <- summary(model)$coefficients[-1, "Estimate"]

  # Store
  pi_unadj_coefficients_clinical[[predictor]] <- coef
}

# Convert the list to a dataframe for better readability
pi_coefficients_df_clinical <- data.frame(
  predictor =
    names(pi_unadj_coefficients_clinical), 
  Coefficient =
    unlist(pi_unadj_coefficients_clinical)
)
```

### PI Clinical + Biomarkers

```{r}
# select the predictor names for the stable set
top10_pi_clbm = stable_sets_clbm %>% filter(health_measure == "pi") %>% select(predictor) %>% 
  unlist() %>% unname()

# Subset model matrix to those in the top 10
X_selected <- mm_pi_clbm$X[, top10_pi_clbm]

# new dataset -- filter the analysis dataset to those with biomarkers and choose appropriate predictors 
pi_data_clbm = analysis_dataset %>% 
filter(!is.na(IFNy)) %>% 
  select(subject_id, pi_traj_cont, illness6) %>% 
cbind(X_selected) %>% 
  na.omit()
```

```{r}
# running the unadjusted regression models

# Initialize a list to store coefficients
pi_unadj_coefficients_clbm <- list()

# Loop through each covariate
for (predictor in top10_pi_clbm) {
  formula <- as.formula(paste0("pi_traj_cont ~ `", predictor, "`"))
  
  # Fit the linear model
  model <- lm(formula, data = pi_data_clbm)
  
  # Extract the coefficient for the covariate
  coef <- summary(model)$coefficients[-1, "Estimate"]

  # Store
  pi_unadj_coefficients_clbm[[predictor]] <- coef
}

# Convert the list to a dataframe for better readability
pi_coefficients_df_clbm <- data.frame(
  predictor = names(pi_unadj_coefficients_clbm), 
  Coefficient = unlist(pi_unadj_coefficients_clbm)
)
```


## Adjusted 

### PI Biomarkers

```{r}
# run the adjusted model with all variables in the stable set, remove the ids 
adjusted_model = lm(pi_traj_cont ~ ., data = pi_data_bm[,-1])

# Extract the coefficients for the covariates
## removing those associated with the intercept and the illness variable
adjusted_coefficients <- summary(adjusted_model)$
  coefficients[-c(1:3),"Estimate"]

# Add the adjusted coefficients to the dataframe
pi_coefficients_df_bm$Adjusted_Coefficient <- adjusted_coefficients
```


### PI Clinical

```{r}
# run the adjusted model with all variables in the stable set, remove the ids 
adjusted_model = lm(pi_traj_cont ~ ., data =
                      pi_data_clinical[,-1])

# Extract the coefficients for the covariates
## removing those associated with the intercept and the illness variable
adjusted_coefficients <- summary(adjusted_model)$
  coefficients[-c(1:3), "Estimate"]

# Add the adjusted coefficients to the dataframe
pi_coefficients_df_clinical$Adjusted_Coefficient <- adjusted_coefficients
```

### PI Clinical + Biomarkers

```{r}
# run the adjusted model with all variables in the stable set, remove the ids 
adjusted_model = lm(pi_traj_cont ~ ., data = pi_data_clbm[,-1])

# Extract the coefficients for the covariates
## removing those associated with the intercept and the illness variable
adjusted_coefficients <- summary(adjusted_model)$
  coefficients[-c(1:3), "Estimate"]

# Add the adjusted coefficients to the dataframe
pi_coefficients_df_clbm$Adjusted_Coefficient <- adjusted_coefficients
```


# Stability Table 

## PI Clinical 

```{r}
# join the coefficients with the stable set metrics
pi_stab_clinical = pi_coefficients_df_clinical %>%
  left_join(stable_sets_clin %>%
              filter(health_measure == "pi") %>% 
              select(predictor, stabscore,
                     median_sp), by = "predictor")

# Rename predictors manually 
pi_stab_clinical$predictor = c("PHQ9", "Diabetes: No", 
                               "Race: Black/AA", 
                               "Trail-Making Test B: Missing", 
                               "Race: More than 1", 
                               "Financial Stress: Enough $ but Little Spare",
                               "Age", "Financial Stress: Difficulty Paying Bills", 
                               "Race: Missing", "Race: Asian", "3-Minute Walk", 
                               "Ethnicity: Missing", "3MS", "Trail-Making Test A",
                               "Psychosocial Resilience Scale: Missing", "15-Item Word List",
                               "3MS Missing", "eGFR", "eGFR: Missing")

# Change columns names for readability
colnames(pi_stab_clinical) = c("Predictor", "Unadjusted Coefficient", 
                               "Adjusted Coefficient", "Stability Score", 
                               "Median Selection Probability")

# Add description column manually
pi_stab_clinical$Description = 
  c("Higher score -> More Depression Symptoms", 
    "n = 110", 
    "n = 19", 
    "n = 6", 
    "n = 2",
    "n = 21",
    "Higher value -> Older", 
    "n = 1",
    "n = 1",
    "n = 2",
    "Higher score -> Better Walking Endurance",
    "n = 1", 
    "Higher score -> Better Cognitive Functioning", 
    "Higher score -> Slower Processing Speed",
    "n = 30", 
    "Higher score -> Better Verbal Memory", 
    "n = 4",
    "Higher score -> Better Kidney Function", 
    "n = 8")

# select needed columns
pi_stab_clinical = pi_stab_clinical %>% select(Predictor, Description, 
                                               `Unadjusted Coefficient`, 
                                               `Adjusted Coefficient`, 
                                               `Stability Score`, 
                                               `Median Selection Probability`)

# Create a flextable
ft <- pi_stab_clinical %>%
  flextable() %>%
  theme_vanilla() %>%
  autofit() %>%
  colformat_double(j = c("Unadjusted Coefficient", 
"Adjusted Coefficient", "Median Selection Probability"), 
digits = 3) %>%
  align(align = "center", part = "all") %>%
  bold(bold = TRUE, part = "header") %>%
  fontsize(size = 10, part = "all")

# Export to Word
# save_as_docx(ft, path = INSERT PATH HERE))
```

## PI Biomarkers 

```{r}
# join the coefficients with the stable set metrics
pi_stab_bm = pi_coefficients_df_bm %>% left_join(stable_sets_bm %>%
filter(health_measure == "pi") %>% 
select(predictor, stabscore, median_sp), by = "predictor")

# Rename predictors manually 
pi_stab_bm$predictor = c("IL27", "IL7", "LBP", "IL5", "Leptin", "IL1Ra",    
                         "IL37", "LIF", "IL15", "IFNa2a", "IFNy", "IL1B", 
                         "TNFRI", "IL8", "Cortisol", "TNFa")

# Change columns names for readability
colnames(pi_stab_bm) = c("Predictor", "Unadjusted Coefficient", 
"Adjusted Coefficient", "Stability Score", "Median Selection Probability")

# select needed columns
pi_stab_bm = pi_stab_bm %>% select(
  Predictor, `Unadjusted Coefficient`, `Adjusted Coefficient`, `Stability Score`, 
  `Median Selection Probability`)

# Create a flextable
ft <- pi_stab_bm %>%
  flextable() %>%
  theme_vanilla() %>%
  autofit() %>%
  colformat_double(j = c("Unadjusted Coefficient", 
"Adjusted Coefficient", "Median Selection Probability"), 
digits = 3) %>%
  align(align = "center", part = "all") %>%
  bold(bold = TRUE, part = "header") %>%
  fontsize(size = 10, part = "all")

# Export to Word
# save_as_docx(ft, path = INSERT PATH HERE)
```

## PI Clinical + Biomarkers

```{r}
# join the coefficients with the stable set metrics
pi_stab_clbm = pi_coefficients_df_clbm %>% 
  left_join(stable_sets_clbm %>% filter(health_measure == "pi") %>% 
              select(predictor, stabscore, median_sp), by = "predictor")

# rename predictors manually
pi_stab_clbm$predictor = c("PHQ9", "Race: Black/AA", "Diabetes: No", 
                           "Trail-Making Test B: Missing", "IL27", "IL7", 
                           "Leptin", "Race: More than 1", "IL1Ra", 
                           "Financial Stress: Enough $ but Little Spare", 
                           "IL5", "LBP", "Trail-Making Test A")

# change names of columns for readability 
colnames(pi_stab_clbm) = c("Predictor", "Unadjusted Coefficient", 
"Adjusted Coefficient", "Stability Score", "Median Selection Probability")

# add decsription column manually
pi_stab_clbm$Description = 
  c("Higher score -> More Depression Symptoms", 
    "n = 19", 
    "n = 110", 
    "n = 6", 
    "",
    "",
    "",
    "n = 2", 
    "",
    "n = 21", 
    "",
    "",
    "Higher score -> Slower Processing Speed")

# select needed columns
pi_stab_clbm = pi_stab_clbm %>% select(
  Predictor, Description, `Unadjusted Coefficient`, `Adjusted Coefficient`, 
  `Stability Score`, `Median Selection Probability`)

# Create a flextable
ft <- pi_stab_clbm %>%
  flextable() %>%
  theme_vanilla() %>%
  autofit() %>%
  colformat_double(j = c("Unadjusted Coefficient", 
"Adjusted Coefficient", "Median Selection Probability"), 
digits = 3) %>%
  align(align = "center", part = "all") %>%
  bold(bold = TRUE, part = "header") %>%
  fontsize(size = 10, part = "all")

# Export to Word
# save_as_docx(ft, path = INSERT PATH HERE)
```

# Stability Heat Map 

## PI Clinical 

```{r}
# set up dataset for heatmap compatability 
heatmap_pi = stable_sets_clin %>% filter(health_measure == "pi") %>%
select(predictor, ends_with("cpss"), ends_with("boot")) 

# rename predictors manually
heatmap_pi$predictor = c("PHQ9", "Diabetes: No", "Race: Black/AA", 
                         "Trail B: Missing", "Race: More than 1", 
                         "Finance: Enough $", "Age", "Finance: Difficulty", 
                         "Race: Missing", "Race: Asian", "3-Minute Walk", 
                         "Ethnicity: Missing", "3MS", "Trail A", 
                         "Psychosocial Resil: Missing", "15-Item Word List", 
                         "3MS Missing", "eGFR", "eGFR: Missing")

# Change names for readability 
heatmap_pi <- heatmap_pi %>%
  rename_with(
    .fn = ~ str_replace(.x, "^q(\\d+)_cpss$", "C\\1")
  ) %>%
  rename_with(
    .fn = ~ str_replace(.x, "^q(\\d+)_boot$", "B\\1")
  )


# Reshape to long format
df_long <- heatmap_pi %>%
  mutate(predictor = factor(predictor, levels = rev(predictor))) %>%
  pivot_longer(
    # adjust if your columns have a different pattern
    cols = starts_with("C") | starts_with("B"), 
    names_to = "configuration",
    values_to = "sp"
  ) %>% 

  group_by(configuration) %>%
  mutate(
    # TRUE if this row’s SP is in the top 10 for this configuration
    top10 = rank(-sp, ties.method = "min") <= 10
  ) %>%
  ungroup()

# relevel factors
df_long$configuration <- factor(df_long$configuration,
                                levels = c("B5", "B10", "B15", "B20", "B25", 
                                           "C5", "C10", "C15", "C20", "C25"))

# Create heatmap
plot_clin = ggplot(df_long, aes(x = configuration, y = predictor)) +
  geom_tile(aes(fill = sp), color = "white") +
  geom_point(data = df_long %>% filter(top10),
             aes(x = configuration, y = predictor),
             shape = 21, color = "black", fill = "yellow", size = 2) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()) +
  labs(
    x = "Configuration",
    y = "Predictor",
    fill = "SP",
    title = "SP Heatmap with Top 10 Indicators"
  )
```

## PI Biomarkers 

```{r}
# set up dataset for heatmap compatability 
heatmap_pi = stable_sets_bm %>% filter(health_measure == "pi") %>%
select(predictor, ends_with("cpss"), ends_with("boot")) 

# rename predictors manually
heatmap_pi$predictor = c("IL27", "IL7", "LBP", "IL5", "Leptin", "IL1Ra",    
                         "IL37", "LIF", "IL15", "IFNa2a", "IFNy", "IL1B", "TNFRI", 
                         "IL8", "Cortisol", "TNFa")

# Change names for readability 
heatmap_pi <- heatmap_pi %>%
  rename_with(
    .fn = ~ str_replace(.x, "^q(\\d+)_cpss$", "C\\1")
  ) %>%
  rename_with(
    .fn = ~ str_replace(.x, "^q(\\d+)_boot$", "B\\1")
  )


# Reshape to long format
df_long <- heatmap_pi %>%
  mutate(predictor = factor(predictor, levels = rev(predictor))) %>%
  pivot_longer(
    # adjust if your columns have a different pattern
    cols = starts_with("C") | starts_with("B"),       
    names_to = "configuration",
    values_to = "sp"
  ) %>% 
  group_by(configuration) %>%
  mutate(
    # TRUE if this row’s SP is in the top 10 for this configuration
    top10 = rank(-sp, ties.method = "min") <= 10
  ) %>%
  ungroup()

# relevel factor
df_long$configuration <- factor(df_long$configuration,
                                levels = c("B5", "B10", "B15", "B20", "B25", 
                                           "C5", "C10", "C15", "C20", "C25"))

# Create heatmap
plot_bm = ggplot(df_long, aes(x = configuration, y = predictor)) +
  geom_tile(aes(fill = sp), color = "white") +
  geom_point(data = df_long %>% filter(top10),
             aes(x = configuration, y = predictor),
             shape = 21, color = "black", fill = "yellow", size = 2) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()) +
  labs(
    x = "Configuration",
    y = "Predictor",
    fill = "SP",
    title = "SP Heatmap with Top 10 Indicators"
  )
```

## PI Clinical + Biomarkers 

```{r}
# set up dataset for heatmap compatability 
heatmap_pi_clbm = stable_sets_clbm %>% filter(health_measure == "pi") %>%
select(predictor, ends_with("cpss"), ends_with("boot")) 

# rename predictors manually
heatmap_pi_clbm$predictor = c("PHQ9", "Race: Black/AA", "Diabetes: No", 
                              "Trail B: Missing", "IL27", "IL7", "Leptin", 
                              "Race: More than 1", "IL1Ra", "Finance: Enough $", 
                              "IL5", "LBP", "Trail A")

# Change names for readability 
heatmap_pi_clbm <- heatmap_pi_clbm %>%
  rename_with(
    .fn = ~ str_replace(.x, "^q(\\d+)_cpss$", "C\\1")
  ) %>%
  rename_with(
    .fn = ~ str_replace(.x, "^q(\\d+)_boot$", "B\\1")
  )


# Reshape to long format
df_long <- heatmap_pi_clbm %>%
  mutate(predictor = factor(predictor, levels = rev(predictor))) %>%
  pivot_longer(
    # adjust if your columns have a different pattern
    cols = starts_with("C") | starts_with("B"),       
    names_to = "configuration",
    values_to = "sp"
  ) %>% 
  group_by(configuration) %>%
  mutate(
    # TRUE if this row’s SP is in the top 10 for this configuration
    top10 = rank(-sp, ties.method = "min") <= 10
  ) %>%
  ungroup()

# relevel factor
df_long$configuration <- factor(df_long$configuration,
                                levels = c("B5", "B10", "B15", "B20", "B25", 
                                           "C5", "C10", "C15", "C20", "C25"))

# Create heatmap
plot_clbm = ggplot(df_long, aes(x = configuration, y = predictor)) +
  geom_tile(aes(fill = sp), color = "white") +
  geom_point(data = df_long %>% filter(top10),
             aes(x = configuration, y = predictor),
             shape = 21, color = "black", fill = "yellow", size = 2) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank()) +
  labs(
    x = "Configuration",
    y = "Predictor",
    fill = "SP",
    title = "SP Heatmap with Top 10 Indicators"
  )
```


## Combine Plots 

```{r, fig.width = 9, fig.height = 8}

scale_sp <- scale_fill_gradient(
  name = "Selection probability",
  low  = "white",
  high = "blue",
  limits = c(0, 1)
)

plot_clin  <- plot_clin  + scale_sp + ggtitle("(a) Clinical predictors only") +  theme(plot.title = element_text(face = "bold"))
plot_bm    <- plot_bm   + scale_sp+ ggtitle("(b) Biomarkers only") +  theme(plot.title = element_text(face = "bold"))
plot_clbm  <- plot_clbm  + scale_sp+ ggtitle("(c) Clinical + biomarkers") +  theme(plot.title = element_text(face = "bold"))


combined_plot <-
  (plot_clin | plot_bm) /
  (plot_clbm | guide_area()) +
  plot_layout(
    guides = "collect",
    widths = c(1, 0.15),  # right column narrow
    heights = c(1, 1)
  ) &
  theme(
    legend.position  = "right",
    legend.direction = "vertical"
  )



combined_plot
```


# Session information

```{r echo = FALSE, results = 'markup'}
sessionInfo()
```


